---
title: "Multiple Linear Regression"
author: "Ben Denis Shaffer"
date: "October 5, 2016"
---

We begin by fitting a linear model based on the `Acetic , H2S , Lactic` covariates to see how they relate to the taste of cheddar.

```{r, echo=F}

library(faraway)
data("cheddar")
dt <- cheddar
attach(dt)

m1 <- lm(taste ~ Acetic + H2S + Lactic)
summary(m1)

```

Next we fit a second model with the `Acetic , H2S` trandformed to their original scale through the `exp` trandfomration appplied to both.

```{r, echo=F}

m2 <- lm(taste ~ exp(Acetic) + exp(H2S) + Lactic)
summary(m2)

```

If we wanted to compare the models we wouldn't use an F-test because the two models are not nested, and the covariates in model two are not a subset or a subspace of the variables in model 1. Instead we use $R^2$ to get a feel for how well the models fit the data, and model judging by $R^2$ model 1 provides a better fit.

```{r}

summary(m1)$r.squared #model 1
summary(m2)$r.squared #model 2

```

If we increase `H2S` by 0.01 the change in taste, all else being equal, is the change times the estimated coef.

```{r, echo=F}

m1$coefficients[3]*0.01

```

To get a percent change in `H2S` when it is transformed by an `exp`, corresponding to a 0.01 addetive change on the original scale we would take the ratio of $exp(H2S + 0.01) / exp(H2S)$ and multiply by 100.

```{r}

(exp(0.01) - 1)*100 #About 1% change

```

To derive the confidence intervals for the first model we compute the standard errors and use t-value assosiated with type I error rate of 5%.

```{r}

confint(m1)

```

Next we plot the confidence region for the `H2S , Lactic` variables which produces a region bounded by an ellipse.


```{r, echo=F}

library(ellipse)
plot(ellipse(cov(H2S, Lactic)), type = "l", col=2, ylab = "SD(Lactic)", xlab = "SD(H2S)", main = "Confidence Region")
points((Lactic - mean(Lactic))/sd(Lactic),(H2S - mean(H2S))/sd(H2S), pch = 20)


```